Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-27 papers/LOPe/coarray-extensions.tex
Finished second reading of programming model.

The transfer of halo memory is necessary after each completion of the do concurrent loop.
Completion of a do concurrent construct indicates that all executing threads (if a threading
model is used by the compiler) have completed and that all memory on the executing subimage is
in a consistent state.

@@

%%where {\tt A} and {\tt Avg} are arrays, but where the {\tt a} argument in the {\tt convolve} definition is declared as a scalar.  The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.  The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.  In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.

\subsection{Memory Management}

@@

@@

>>>>

Note that a transfer of halo memory is necessary after each completion of the do concurrent loop.
This must be done in order for the halo region of a coarray on a given process to be consistent
with the corresponding interior of the coarray on a logically neighboring process.

%% However, recall that with the LOPe programming model only the local element is mutable; thus, the halo region of an array needs update.

@@

\subsection{Execution Symantics and Memory Management}

@@

%% Completion of a do concurrent construct indicates that all executing threads (if a threading model is used by the compiler) have completed and that all memory on the executing subimage is in a consistent state.

%%The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.  The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.  In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.

@@

\input{coarray-comparison}

Note that the use of halo cells is the normal way that large and complex MPI and CAF
programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
language thus providing the compiler information in order to spread computation over more
hardware resources and reduce complexity for the programmer.

### 2015-01-27 papers/LOPe/local.tex
First cut at programming-model (mostly all there) stage.
Finished second reading of programming model.

Emerging hybrid accelerator architectures for high performance computing are often well suited for
usage of a data-parallel programming model.  However, programmers for these emerging architectures
face a steep learning curve that frequently requires learning a new language, e.g., OpenCL or CUDA,
or employing OpenMP compiler directives.  Furthermore, the distributed (and frequently multi-level)
nature of the memory organization of clusters of these architectures provides an additional level of
complexity for the programmer.  This paper presents preliminary work examining how programming with
a local orientation can be employed to provide simple access to accelerator architectures.  A
locally-oriented programming model is especially useful for stencil-based codes or algorithms
requiring the application of convolution kernels.  In this programming model, a programmer codes the
algorithm by modifying \emph{only a single array element}, but has read-only access to a small
sub-array surrounding the given array element, so that a stencil can be applied.  We demonstrate how
a locally-oriented programming model can be adopted as an embedded, domain-specific language using
source-to-source program transformations.

@@

\input{motivation}
\input{programming-model}
\input{appendix-a}

>>>>

Emerging hybrid accelerator architectures for high performance computing are often suited for the
use of a data-parallel programming model.  Unfortunately, programmers for these emerging
architectures face a steep learning curve that frequently requires learning a new language, e.g.,
OpenCL or CUDA, or employing OpenMP compiler directives.  Furthermore, the distributed (and
frequently multi-level) nature of the memory organization of clusters of these architectures
provides an additional level of complexity for the programmer.  This paper presents preliminary work
examining how programming with a local orientation can be employed to provide simple access to
accelerator architectures.  A locally-oriented programming model is especially useful for the
solution of algorithms requiring the application of a convolution kernel.  In this programming
model, a programmer codes the algorithm by modifying \emph{only a single array element} (called the
local element), but has read-only access to a small sub-array surrounding the local element so that
a stencil can be applied.  We demonstrate how a locally-oriented programming model can be adopted as
an embedded, domain-specific language using source-to-source program transformations.

@@

\input{motivation}
\input{programming-model}
\input{coarray-extensions}
\input{appendix-a}

### 2015-01-27 papers/LOPe/motivation.tex
Finished second reading of programming model.

The concepts in this paper were motivated by problems encountered during
the development of PetaVision, a C++ software framework designed for simulating
large networks of spiking neurons in the visual cortex of primates.  PetaVision
was designed to run on massively parallel hardware architectures and a primitive
version of PetaVision achieved over a Petaflop/s of sustained single precision
performance on Roadrunner, the fastest computer in the world for a brief
period of time.

@@

TODO: Give a simple 1D (cartoon) example of connections within a neural network in
Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
the number of synaptic connections, and performance.  Reference SC Gorden Bell
paper from IBM.

>>>>

The concepts in this paper were motivated by problems encountered during
the development of PetaVision, a C++ software framework designed for simulating
large networks of spiking neurons in the visual cortex of primates.  PetaVision
was designed to run on massively parallel hardware architectures and a primitive
version of PetaVision achieved over a Petaflop of sustained single precision
performance on Roadrunner, the fastest computer in the world for a brief
period of time.

@@

TODO: If space allows, give a simple 1D example of connections within a neural network in
Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
the number of synaptic connections, and performance.  Reference SC Gorden Bell
paper from IBM.

TODO: Explain difficulties encountered.

### 2015-01-27 papers/LOPe/programming-model.tex
Finished second reading of programming model.

The LOPe (for Locally-Oriented Programming extensions) programming model restricts the programmer ---
when implementing a stencil algorithm --- to a local view of the array index space.  Within a LOPe
function, only the local array element (plus a small halo region surrounding the local element) is
visible to the programmer.  This serves to reduce complexity by removing all boundary conditions and
processor topology information from the implementation of the algorithm.

@@

There are only a few syntax additions required for LOPe.  In the code examples describing these
additions (and throughout this paper as well), DSL syntax additions are highlighted by the usage of
capitalization for all new keywords.

\subsubsection{HALO.}

The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
an array boundary, so that the array may be referenced using indicies that fall ``outside'' of the
domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
exploit this knowledge.  For example, a halo region can be explicitly provided in a declaration
statement of the form,
\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}
This statement indicates that array \texttt{A} is one dimensional, will be allocated later when it
is provided with a specific size, and has a halo region of 1 surrounding the array on both sides.
The notation \texttt{M:*:N} is to be read as describing a halo region of size \texttt{M} to the
left, size \texttt{N} to the right, and with an arbitrary size in the middle (denoted by the $*$
symbol).  When used to describe a formal parameter of a function, the \texttt{HALO} keyword may
``assumed'' (from the actual argument by the compiler) as is the following statement,
\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}
which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
whose size is ``assumed'' by the compiler from the explicit halo size of the actual array argument
(provided at the calling site of the function).  In LOPe, there is no need in this instance for a
repetitive \texttt{dimension(:,:)} specification as it is inferred by the \texttt{HALO}
specification.

\subsubsection{CONCURRENT.}

The second keyword employed by LOPe is \texttt{CONCURRENT}.  The \texttt{concurrent} keyword already
exists in the form of a \texttt{do concurrent} loop construct (a Fortran 2008 feature).  By using
\texttt{do concurrent}, the programmer asserts that specific iterations of the loop body may be
executed by the compiler in \emph{any order;} even \emph{concurrently.}  For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  A function with the \texttt{pure} and \texttt{CONCURRENT} attributes may be called from
with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
\texttt{(i,j)} iterate of the loop.  An example of this usage will be given later in the text.

\subsubsection{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
array element and only the local element may be modified.  The default zero-based indexing scheme
for the local array element is somewhat different that normal Fortran notation in that by default,
array indices start a 1.  For LOPe, this default convention is modified; indices for arrays declared
with the \texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in
one dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local) and
\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local).  The geometric positioning of the
array elements can be seen by the arrangement of the expressions on the right-hand side of the
\texttt{Laplace} example.

\input{coarray-comparison}

Note that the use of halo cells is the normal way that large and complex MPI and CAF
programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
language thus providing the compiler information in order to spread computation over more
hardware resources and reduce complexity for the programmer.

>>>>

The LOPe programming model (for Locally-Oriented Programming extensions) restricts the programmer to
a local view of the index space of an array.  Within a LOPe function, only a single array element (called
the local element) is mutable.  In addition, a small halo region surrounding the local element is
visible to the programmer, but this region is immutable.  Restricting the programmer to a local
index space serves to reduce complexity by removing all boundary-value and processor-topology
information from the implementation of the algorithm.

@@

There are only a few syntax additions required for a LOPe program.  These additions include syntax for
describing halo regions and concurrent procedures.  In all of the code examples that follow, DSL
syntax additions are highlighted by the usage of capitalization for keywords that are either new or
that acquire new usage.

\subsubsection{Halo regions.}

The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
an array boundary, so that an array may be referenced using indices that fall ``outside'' of the
domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}
This statement indicates that \texttt{A} is a rank one array, will be allocated later (when it will
be provided with a specific size), and has a halo region of one element surrounding the array on
either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of size
\texttt{M} to the left, size \texttt{N} to the right, and with an arbitrary size in the middle
(denoted by the $*$ symbol).  When used to describe a formal parameter of a function, the size may
be ``assumed'' (denoted by a ``:'' symbol substituted in place of an explicit size) as is the
following,
\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}
which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
whose size is obtained by the compiler from the explicit halo size of the actual array argument
provided at the calling site of the function.  In LOPe, there is no need (in this instance) for a
repetitive \texttt{dimension(:,:)} specification as it is inferred from the \texttt{HALO}
specification.

\subsubsection{Concurrent functions.}

The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
of a \texttt{do} \texttt{concurrent} loop construct.  By using this construct, the programmer
asserts that specific iterations of the loop body may be executed by the compiler in \emph{any
  order;} even \emph{concurrently.}
%%LOPe allows the \texttt{concurrent} attribute to be attached to a function declaration so that the function can be called from within a concurrent loop without fear of race conditions.
For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  LOPe allows a function with the \texttt{pure} and \texttt{concurrent} attributes to be
called from within a \texttt{do} \texttt{concurrent} loop body.  One should imagine that this
function is called \emph{for each} \texttt{(i,j)} index of the interior of the array \texttt{U}.
(An example of this usage will be given later in the text.)  Note that this usage introduces a race
condition as new values of elements of \texttt{U} are created on the left-hand side of the
assignment statement that may use \emph{new or old} values of \texttt{U} on the right-hand side.
LOPe requires the compiler to guarantee that race conditions won't occur by using, e.g.,
double-buffering techniques as needed.

\subsubsection{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
array element and only the local element may be modified.  This zero-based indexing for the
local-array element is different than conventional Fortran, where by default, array indices start at
1.  For LOPe, this default convention is modified; indices for arrays declared with the
\texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in one
dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local, respectively) and
\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local, respectively).  The geometric
positioning of the array elements can be seen by examining the arrangement of the expressions on the
right-hand side of the \texttt{Laplace} example.
